---
title: "US Covid Data spatial temporal clustering"
author: "Ruiman Zhong"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. sfclust - Real Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction


Disease clustering, particularly functional disease clustering, plays a crucial role in identifying patterns that help in understanding and managing health crises. By grouping regions with similar disease trajectories, it becomes possible to detect early signs of outbreaks and allocate resources effectively. Functional clustering, in particular, captures changes over time, adding an additional layer of insight into disease progression.

The `sfclust` package provides tools for Bayesian spatial functional clustering, which is especially useful in detecting clusters of neighboring regions with similar disease risks. These clusters can help uncover inequalities and guide the distribution of health resources based on risk levels.

Compared to traditional spatial clustering methods, `sfclust` offers greater flexibility by allowing each cluster to have its own unique parameters, offering a more adaptable model. Additionally, the number of clusters is not restricted by their size or shape, and spatial contiguity is ensured by the initial full graph structure. The method also supports the analysis of non-Gaussian data, with latent functions represented by different processes, making it suitable for a wide range of applications.

This vignette demonstrates how to use the `sfclust` package to perform Bayesian spatial functional clustering on COVID-19 data from the United States, illustrating its application to disease mapping.

# Loading Necessary Packages

We begin by loading the required packages. These packages will assist with data manipulation, visualization, and running the `sfclust` algorithm.

```{r include=FALSE}
packages <- c("stars", "deldir", "fields", "igraph", 
              "ggplot2",  "grid", "class", 'sf', 'sfclust','ggraph')
lapply(packages, require, character.only = TRUE)
```
# Loading and Preparing the Data

We will use a pre-processed dataset that contains COVID-19 case counts (`Y`), population data (`N`), and a `map` object representing the spatial layout of the regions. The data covers both spatial and temporal dimensions, focusing on the weekly COVID-19 cases across 49 states in the U.S (Hawaii and Alaska are not included) during the 2020 (51 weeks in total). Each state’s infection data is captured over time to identify how the pandemic evolved geographically. 

The dataset is stored in `.rds` format, pre-processed for this analysis. It is a `stars` object, including response `Y`, expected cases `E`, and the total population `N`.
```{r warning=FALSE}
data("us_covid")
map <- st_as_sf(st_get_dimension_values(us_covid, "space"))
st_geometry(map) <- "geometry"
```

# Running the Clustering Algorithm

We'll now use the `sfclust()` function to perform clustering based on the preprocessed data. This step includes building a Minimum Spanning Tree (MST) and running the Bayesian spatial functional clustering algorithm.
The function `initial_cluster() ` is to build a initial partition parameters (Graph, MST, cluster). We could generate the initial graph by adjacency matrix. Other possible ways to construct graph are provided in another vignette. The number of the initial cluster is set by `nclust`. If the `weights` = NULL, means the weights of every egdge in the initial graph is random generated from U[0,1].

```{r warning=FALSE, fig.width=6, fig.height=4}
# Initialize the graph data using Minimum Spanning Tree (MST)
geodata <- initial_cluster (map, nclust = 10)
coords <- st_coordinates(st_centroid(map))
plot(geodata$graph, layout = coords)
```

We could also check the initial MST and cluster memberships, by

```{r fig.width=6, fig.height=4}
plot(geodata$mst, layout = coords)
```


The plot illustrates that we have 10 initial clusters. 


Next, we will define the formula for the Bayesian spatial functional with-cluster model. Here we use AR 1 model, where `Yk` is the response of each cluster (Covid cases in the US), and `Xk` is the random effect. If covariates are provided, we could include them in the fixed effect. The package supports common model that `R-INLA` supports. For more information, please check `R-INLA`.
$$Y_{it} \mid \mu_{it}, \boldsymbol{\theta}_{T} ~\stackrel{ind}{\sim}~ \text{Poisson}(E_{i} \times \mu_{it})$$
$$\log(\mu_{it}) = \eta_{it} = \alpha_{c_i} + f_{c_i,t} + \epsilon_{it}, ~\text{where}~ \epsilon_{it} \mid \boldsymbol{\theta}_{T} {\sim} N(0, 1/\tau_{c_i}^2),$$
$$f_{c,1} \mid v_{c}, \rho_{c}, \boldsymbol{\theta}_{T} \sim N(0, v_{c}^{-1}(1-\rho_{c}^2)^{-1}), ~~\text{for}~~ c = 1, \dots, C,$$
$$f_{c, t} = \rho_{c} f_{c, t-1} + \varepsilon_{c,t} ~\text{for}~ t = 2, \dots, T~\text{where}~ \varepsilon_{c,t} \mid \boldsymbol{\theta}_{T} \sim N(0, v_{c}^{-1})$$
```{r}
# Define the formula for the Bayesian spatial functional model
formula <- Y ~ 1+ f(idt, model = "ar", order = 1) + f(id, model = "iid")
```
 
 `sfclust()` is the main functions to implement Bayesian spatial functional clustering based on bayesian spanning tree.  `Y` is the response, here is the Covid cases in the US. `idt` is the index of the time, created by our function. 
```{r}
data <- preprocess_data_each(us_covid, 3, geodata$cluster, time_var = "time")
head(data)
```
 
 We fit a `poisson` likelihood with the offset  `N` = E, where E is the expected cases. Hyperparameter `q` describe the penalty of the number of the final cluster, ranging from 0 to 1. We don't need any correction when using ar model. We will run `nter` = 1000 MCMC chain, without burn-in procedure. 

 
 The results of the function is saved in the path_res.rds. Please define a valid address and name before running the function. The function also return the results. 
 
```{r eval=FALSE}
# define the path to save the final results, runs for one hour
path <- '~/Documents/project3'
path_res <- file.path(path, "data", "result","US_cluster.rds")
# Run the bsfc algorithm with the defined parameters
result <- sfclust(us_covid, graphdata = geodata, N_var = 'E', formula = formula, 
     family = "poisson", q = 0.5, 
     correction = FALSE, niter = 1000, burnin = 0, 
     thin = 1, path_save = path_res)
```

# Analyzing the Results

After running the model, we can analyze the results, including the final model and cluster assignments. There are five components in total in the results: the each cluster membership at each step, the related marginal likelihood at each step, the MST at each step, and the final estimated INLA model results.

Let's start by loading the initial results and analyzing the MCMC chain to assess convergence.
```{r include=FALSE}
path <- '~/Documents/project3'
path_res <- file.path(path, "data", "result","US_cluster.rds")
# Load the results from the saved file
result <- readRDS(path_res)
```
```{r fig.width=6, fig.height=4}
plot(result, map)
```


The plot above shows the marginal likelihoods across iterations. If the plot indicates that the chain has not converged, we can continue the MCMC process.

# Continuing the MCMC Process
Next, we'll use the continue_bsfc function to extend the MCMC process by adding 2000 more iterations.
```{r include=FALSE}
# Define the additional MCMC settings
path_res <- file.path(path, "data", "result","US_cluster2.rds")
```
Run the continuation of the BSFC process
```{r eval=FALSE}
niter_additional <- 4000
# the function runs for one hour
result_continue <- continue_sfclust(result,
  us_covid, graph = geodata$graph,
  formula = formula, family = "poisson", q = 0.5,
  correction = FALSE, niter = niter_additional, burnin = 0, thin = 1,time_var = "time", N_var = "E",
  path_save = path_res
)
```
# Analyzing the Continued Results

After running the additional MCMC iterations, we can load and analyze the updated results to check for convergence.
```{r, fig.width=6, fig.height=4}
result_continue <- readRDS(path_res)  # Load the continued results into the workspace
mlik <- c(result$log_mlike, result_continue$log_mlike)
# Plot the updated marginal likelihoods
plot(mlik, type = "l", main = "Marginal Likelihoods After Continued MCMC Run",
     xlab = "Iteration", ylab = "Marginal Likelihood")
```

Print the final cluster_membership and plot.

```{r fig.width=6, fig.height=4}
clust_res <- result_continue$cluster[length(result_continue$mst),]
clust_res
plot(result_continue, map)
```

### Check the fitted curves and time series of each cluster
```{r warning=FALSE, fig.width=6, fig.height=4}
library(dplyr)
start_date <- as.Date('2020-01-20')
end_date <- as.Date('2021-01-04')

# Generate a sequence of weekly dates
time <- seq.Date(from = start_date, to = end_date, by = "week")
final_model = result_continue$model

level <- as.numeric(levels(as.factor(clust_res)))
  ydf = setNames(as.data.frame(Y / E), 1:ns) |>
    mutate(time = time) |>
    tidyr::pivot_longer(1:ns, names_to = "region", names_transform = as.numeric) |>
    mutate(cluster = clust_res[region])
  cluster2 = rep(level, table(clust_res))
  preddf = purrr::map(level, function(x) final_model[[x]]$summary.fitted.values$mean / exp(final_model[[x]]$summary.random[['id']]$mean)) %>%
    purrr::map(~ matrix(., nrow = length(time))) %>%
    do.call(cbind, .) |>
    as.data.frame() |>
    setNames(1:49) |>
    mutate(time =  time) |>
    tidyr::pivot_longer(1:49, names_to = "newregion", names_transform = as.numeric) |>
    mutate(cluster = factor(cluster2[newregion]))
  # Define the custom labeller function
  custom_labeller <- function(cluster) {
    return(paste("Cluster", cluster))
  }

  # Apply the custom labeller to the facets
  p <- ggplot(ydf) +
    geom_line(aes(x = time, y = value, group = region), color = "gray", linewidth = 0.2) +
    geom_line(data = preddf, aes(x = time, y = value, group = newregion),
              color = "red", linewidth = 0.3) +
    facet_wrap(~ cluster, ncol = 4, scales = "free_y", labeller = labeller(cluster = as_labeller(custom_labeller))) +
    theme_bw() +
    theme(legend.position = "none", legend.text = element_text(size = 7)) +
    labs(color = "Cluster", y = "Relative Risk", x = "Time")

custom_breaks <- seq.Date(from = start_date, to = end_date, by = "8 weeks")

g = p  +
  scale_x_date(date_labels = "%b %d", breaks = custom_breaks) +  # Format x-axis labels to show month and day
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
g
```



